Modelling temporal dependence and multilevel structure using hidden Markov models

### 2020-07-22 Report/Sections/appendix.tex
Updates from Overleaf

@@

From here it is straightforward to calculate both $Z^{*(1)}$ and $Z^{*(2)}$. We pick $\tilde{\omega} = 10$ periods per window, or 5 hertz. To find the distribution of $Z^{*(1)}_{t,t^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right)$, use Equation (\ref{eqn:y_hat0}):
%
$$\left(Z^{*(1)}_{t,t^*} | X^*_{t,t^*} = i^* \right) = \left(\mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right) | X^*_{t,t^*} = i^* \right) \sim \mathcal{N} \left(\phi^{*(i^*)} * Z^{*(1)}_{t,t^*-1}, \sigma^{*(i^*)} \right)$$
%
$Z^{*(2)}_{t,t^*}$ is the sum of gamma-distributed random variables with the same scale parameter, so the distribution of $Z^{*(2)}_{t,t^*}$ is also a gamma distribution:
%
$$Z^{*(2)}_{t,t^*} = \sum_{k=1}^{10} b^{(k)}_{t,t^*}$$
$$\left(Z^{*(2)}_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha = \sum_{k=1}^{10} 5/k^2 = 10.10, \beta = 1.00 \right)$$
$$\left(Z^{*(2)}_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha = 300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , \beta = 1.00 \right)$$
\end{proof}

>>>>

Michelot \& Blackwell (2019) model the movement of an animal as the solution to the following stochastic differential equation, which they refer to as a state-switching Ornstein-Uhlenbeck process:
%
\begin{equation}
    \label{eqn:SDE}
    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
\end{equation}
%
where $X_t$ is some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process. Unlike HMMs, $t \in \mathbb{R}$ indexes a continuous time process and is therefore not necessarily an integer. If the behavioral state $X_t$ is known and does not change between observations, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
%
\begin{align}
    Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
    \label{eqn:OU_sol}
\end{align}
%
where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$

\begin{theorem}{Theorem 1.}{}%
If the following conditions are met:
\begin{enumerate}
    \item The hidden behavioural process $X$ from Equation \ref{eqn:SDE} follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$,

    \item Observations of the SDE from Equation 2 are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
\end{enumerate}
then the observations $Y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:

\begin{align}
\mu^{(i)} = \gamma^{(i)}, \qquad \sigma^{(i)} = \sqrt{\frac{\omega^{(i)^2}}{2\beta^{(i)}} (1-e^{-2\beta^{(i)}\Delta t})}, \qquad \phi^{(i)} = e^{-\beta^{(i)}\Delta t} \label{eqn:CarHMM_to_OU}
\end{align}

\end{theorem}

@@

From here it is straightforward to calculate both $Z^{*(1)}$ and $Z^{*(2)}$. We pick $\tilde{\omega} = 10$ periods per window, or 5 hertz. To find the distribution of $Z^{*(1)}_{t,t^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right)$, use Equation \ref{eqn:y_hat0}:
%
$$\left(Z^{*(1)}_{t,t^*} | X^*_{t,t^*} = i^* \right) = \left(\mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right) | X^*_{t,t^*} = i^* \right) \sim \mathcal{N} \left(\phi^{*(i^*)} * Z^{*(1)}_{t,t^*-1}, \sigma^{*(i^*)} \right)$$
%
$Z^{*(2)}_{t,t^*}$ is the sum of gamma-distributed random variables with the same scale parameter, so the distribution of $Z^{*(2)}_{t,t^*}$ is also a gamma distribution:
%
$$Z^{*(2)}_{t,t^*} = \sum_{k=1}^{10} b^{(k)}_{t,t^*}$$
$$\left(Z^{*(2)}_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha = \sum_{k=1}^{10} 5/k^2 = 10.10, \beta = 1.00 \right)$$
$$\left(Z^{*(2)}_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha = 300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , \beta = 1.00 \right)$$
\end{proof}

### 2020-07-22 Report/Sections/discussion.tex
Updates from Overleaf

We presented a collection of HMM models which can be combined together to form increasingly complex hierarchical models to match the complexity of particular problems faced by researchers. This flexible framework can be used to deal with complicated dependence structures within time-series data.

Traditional HMMs can be used to model a state-switching process with conditionally independent observations and Markovian dynamics when conditioned of the hidden state. However, many real-world processes are more complicated than this and require more complex models.

@@

Although the CarHMM can incorporate auto-correlation into the structure of an HMM, it can break down when observations are taken at irregular time intervals. A common solution to this issue is to use a continuous-time method such as the state-switching OU processes described by Michelot \& Blackwell (2019). Most continuous time models require relatively slow MCMC algorithms to perform inference, and as a result are not easily incorporated into the HHMM structure. However, we show in Theorem 1 that certain continuous-time methods are equivalent to an CarHMM under certain conditions (see appendix for proof).

For simultaneous observed processes taking place at different time scales, the HHMM \citep{Barajas:2017,Adam:2019} utilizes hierarchical structures to jointly model both as HMMs. In particular, each hidden state of the coarse-scale HMM is assumed to emit both an observation $Y_t$ as well as another fine-scale HMM with hidden states $X^*_t$ and observations $Y^*_t$.

For processes with very high sampling frequencies and/or with intricate fine-scale structure, it is possible to generalize the HHMM such that the fine-scale model can be any model which admits an easy-to-calculate likelihood. For example, if the sampling rate of the fine-scale process is very high, then the fine-scale model can be described by a simple probability distribution over the summary statistics of a moving window of observations.

Combining these models together should be done with care, as it is important to balance the need to effectively capture the process in question with the need to avoiding over-fitting and slow parameter estimation.

@@

The simulation study also shows that the observed Fisher information serves as a suitable approximation for the standard errors of parameter estimates in most cases. One notable exception is that the standard errors of the probability transition matrix estimates ($\hat \Gamma$ and $\hat \Gamma^*$) tend to be overestimated by the observed Fisher information.

Finally, we used the CarHHMM-DFT to model the behavior of a killer whale off the coast of British Columbia, Canada. The CarHHMM-DFT was able to simultaneously distinguish three distinct sub-dive behaviors and two dive types. The DFT component proved useful in determining the sub-dive behaviour of the whale, as the mean of the emission distribution of $Z^{*(2)}$ for each sub-dive state was separated by an order of magnitude. Finally, the estimated auto-correlation parameter for $\mathbf{Z}^{*(1)}$, $\phi^*$, was above 0.5 for every dimension and sub-dive type, providing evidence that the conditionally auto-regressive component of the CarHHMM-DFT resulted in a better fit to the data. The introduction of the parameter $\phi^*$ also allows $\mathbf{Z}^{*(1)}$ to be interpreted as a state-switching OU process (see appendix).

@@

>>>>

Traditional HMMs can be used to model a state-switching process with conditionally independent observations and Markovian dynamics when conditioned of the hidden state. However, many real-world processes require more complex models. Here, we presented a flexible framework that can account for complicated dependence structures within time-series data. Our framework is a collection of HMM models which can be combined together to form increasingly complex hierarchical models to match the complexity of particular problems faced by researchers.

@@

Although the CarHMM can incorporate auto-correlation into the structure of an HMM, it can break down when observations are taken at irregular time intervals. A common solution to this issue is to use a continuous-time method such as the state-switching OU processes described by \citet{Michelot:2019}. Most continuous time models require relatively slow MCMC algorithms to perform inference, and as a result are not easily incorporated into the HHMM structure. However, we show in Theorem 1 (see appendix) that the state-switching OU process is equivalent to an CarHMM under certain conditions. This implies that the parameter estimates from the CarHMM are interpretable in the context of a stochastic differential equation.

For simultaneous observed processes taking place at different time scales, the HHMM \citep{Barajas:2017,Adam:2019} utilizes hierarchical structures to jointly model both as HMMs. In particular, each hidden state of the coarse-scale HMM is assumed to emit both an observation $Y_t$ as well as another fine-scale HMM with hidden states $X^*_t$ and observations $Y^*_t$. \citet{Adam:2019} utilize the HHMM to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California. However, they run into clear violations of conditional independence and apparent periodic behavior when plotting the auto-correlation of acceleration as a function of lag, implying the need to account for complicated dependence structures within fine-scale processes with high sampling frequencies.

%For processes such as these with very high sampling frequencies and/or intricate fine-scale structure, 
To deal with these issues, it is possible to generalize the HHMM such that the fine-scale model can be any model which admits an easy-to-calculate likelihood. For example, if the sampling rate of the fine-scale process is very high and clearly period, then the fine-scale model can be described by a simple probability distribution over the summary statistics of a moving window of observations.

In particular, we use $Z^{*(1)}$ and $Z^{*(2)}$ based on the DFT as defined previously. These summary statistics prove to be vital when decoding the behavior of a killer whale. This is evidenced by both our simulation study, where the sub-dive decoding accuracy is nearly perfect when $Z^{*(2)}$ is accounted for, and the case study, where the CarHHMM without DFT visually fails to pick up fluking behavior in the killer whale.

While we combine these models together effectively for our case study, in general this practice should be done with care. It is important to balance the need to effectively capture the process in question with the need to avoid over-fitting and slow parameter estimation.

@@

The simulation study also shows that the observed Fisher information serves as a suitable approximation for the standard errors of parameter estimates in most cases. One notable exception is the HHMM-DFT, which ignores auto-correlation and therefore underestimates the standard error of the parameters associated with acceleration. This can be dangerous in the field of ecology, as overconfident parameter estimates can result in overconfident conservation decisions. While the point estimates are generally good in this case study, this may not be the case in more general settings and for shorter time series in particular.

%We used the CarHHMM-DFT to model the behavior of a killer whale off the coast of British Columbia, Canada. The CarHHMM-DFT was able to simultaneously distinguish three distinct sub-dive behaviors and two dive types. The DFT component proved useful in determining the sub-dive behaviour of the whale, as the mean of the emission distribution of $Z^{*(2)}$ for each sub-dive state was separated by an order of magnitude. Finally, the estimated auto-correlation parameter for $\mathbf{Z}^{*(1)}$, $\phi^*$, was above 0.5 for every dimension and sub-dive type, providing evidence that the conditionally auto-regressive component of the CarHHMM-DFT resulted in a better fit to the data. The introduction of the parameter $\phi^*$ also allows $\mathbf{Z}^{*(1)}$ to be interpreted as a state-switching OU process (see appendix).

@@

In total, this work provides a flexible framework to model functional time-series data using HMM models which are easy to fit using maximum likelihood methods. The model of interest can be easily adjusted to match the complexity of the problem at hand. Researchers can use this method to build model and perform inference in any setting which involves state-switching fine-scale processes, including finance and medicine.

### 2020-07-22 Report/Sections/intro.tex
Updates from Overleaf

Many statistical techniques have been developed and applied to these types of longitudinal or functional data.  
Some techniques, such as time series, are highly parametric. 
While these parametric techniques do not allow the data to ``speak for itself", the ensuing data analysis results typically are readily interpretable, providing important subject area insights.
At the other extreme are black box type methods of machine learning.  
While these methods are incredibly successful at predicting how processes evolve, they provide little structured insight into the driving mechanisms of the process.
Functional Data Analysis (FDA) techniques lie somewhere in between these extremes: the output of an analysis provides visual tools for understanding processes and, often, some dimension reduction to allow for easier exploration.  
The development of structured FDA techniques incorporating more complex dependence structures has become a topic of recent interest  (see, e.g., Crainiceanu. and references therein).

Our goal is to develop movement analysis methods that are flexible but also use subject area knowledge to provide insight into curve data with complex structure.
We  model complexities arising  from curve to curve variation, along with the complexities of within-curve fine-scale characteristics and temporal dependence,
combining existing methods in novel ways.
We enforce temporal structure via Markov processes.  
While our methodology has general applicability, [rephrase - awkward] we  showcase our ideas to study the underlying behaviour indicated by the complex dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.  
For more information about the data, see Section \ref{sec:case_study}.
 Figure \ref{fig:data} shows the data we will analyze.
   The top panel shows the depth of the whale below the surface, and we clearly see dives of various time durations and maximum depths. 
The bottom three panels show the acceleration in each of three orthogonal directions,  as recorded by an accelerometer,   
with axes oriented relative to the whale's forward movement.
Here, we see spikes in acceleration and, in contrast, periods of fairly constant acceleration.
Figure {\ref{fig:labeled_dives}} gives a more detailed view, showing a segment of the data containing six dives of at least two different types.
The plot shows the magnitude of the acceleration and from this, we can see fine-scale behaviour within each dive which seems to indicate some fine scale short-term periodicity. 
We seek to understand both the coarse scale behaviour (e.g.~the dive type) and the fine-scale, within-dive behaviour.  
 Such an analysis helps biologists to understand the foraging behaviour and feeding success of killer whales.

@@

From the plotted killer whale data in Figure {\ref{fig:labeled_dives}}, we see complexities within each dive that would be difficult to model as a realization of a simple stochastic process. 
In FDA, within curve structure is modelled by a generic smooth mean function and covariance function.
Alternatively, this structure is sometimes generated via random regression.
However, these classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in Figure 9.  
We model the within curve dependence by combining a hidden Markov model of within-curve behaviour combined with an autoregressive-type dependence.
Finally, for the fine-scale behaviour, we use summary information that reflects the application area's meaning for the curve.  Specifically, for the dive data, we use summaries of Fourier transforms to account for sudden and extreme ``wiggliness'' within some segments of a curve. 
The distribution of this ``wiggliness'' depends on the hiddent state.

@@

Our work relies on work in the vast field of methodology for the analysis of animal movement
\citep{Hooten:2017}.
One of the most prevalent techniques in the literature of late is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \citep{Patterson:2017}. 
%
%Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another conditioned on the underlying behavioral process.
 %However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
 %For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
 Several publications have dealt with this issue in the past, including the hidden movement Markov model (HMMM) \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model (CarHMM) \citep{Lawler:2019}. 
 The CarHMM in particular explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.
 %while maintaining the structure needed to run the forward algorithm. 
 %It also only adds one additional parameter per possible hidden state.

Animal movement modelers have recognized that several simultaneous behavioral processes may occur at different time scales. 
One approach to the analysis of this type of data is to use a hierarchical hidden Markov model (HHMM) \citep{Barajas:2017} \citep{Adam:2019}. 
HHMMs model the entire time series in question as a nested structure of hidden Markov models (HMM) where each HMM corresponds to one behavioral process.

At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by traditional HMMs, CarHMMs, or HHMMs. Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}. One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation. Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals. However, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.

For periodic behavior in particular, one way to avoid the use of continuous time models is to use signal processing techniques such as the Fourier transform on the raw data. 
The advantages of using Fourier analysis within an HMM has been recently demonstrated in the context of describing daily behavioral cycles of marine mammals \citep{Heerah:2017}.
 In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017} and specifically fluking \citep{Shorter:2017} from accelerometer data. 
 Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appears a promising simple approach to account for additional correlation in data that is cyclical in nature.

We consider several models: the classical hierarchical hidden Markov model (HHMM) \citep{Barajas:2017}, the CarHMM model of \citep{Lawler:2019}, and our new model that blends the two (CarHHMM). 
The HHMM consists of a coarse-scale process and a fine-scale process, with standard HMMs modelling both the coarse- and fine-scale processes. 
The CarHMM modifies the usual HMM by modelling additional dependence into the sequence of observed data. 
In our combination of the HHMM and the CarHHMM models, we keep the usual Markov structure for the coarse-scale process, but use the CarHMM model for the fine-scale process.

>>>>

Many statistical techniques have been developed and applied to these types of longitudinal or functional data.  
Some techniques, such as time series, are highly parametric. 
While these parametric techniques do not allow the data to ``speak for itself", the ensuing data analysis results typically are readily interpretable, providing important subject area insights.
At the other extreme are black box type methods of machine learning.  
While these methods are incredibly successful at predicting how processes evolve, they provide little structured insight into the driving mechanisms of the process.
Functional Data Analysis (FDA) techniques lie somewhere in between these extremes: the output of an analysis provides visual tools for understanding processes and, often, some dimension reduction to allow for easier exploration.  
The development of structured FDA techniques which incorporate more complex dependence structures has become a topic of recent interest  (see, e.g., Crainiceanu. and references therein).

Our goal is to develop movement analysis methods that are flexible but also use subject area knowledge to provide insight into curve data with complex structure.
We  model complexities arising  from curve to curve variation, along with the complexities of within-curve fine-scale characteristics and temporal dependence,
combining existing methods in novel ways.
We enforce temporal structure via Markov processes.  
While our methodology is generally applicable, [rephrase - awkward] we  showcase our ideas to study the underlying behaviour indicated by the complex dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.  
For more information about the data, see Section \ref{sec:case_study}.
Figure \ref{fig:data} shows the data we will analyze.
The top panel shows the depth of the whale below the surface, and we clearly see dives of various time durations and maximum depths. 
The bottom three panels show the acceleration in each of three orthogonal directions, as recorded by an accelerometer,   
with axes oriented relative to the whale's forward movement.
Here, we see spikes in acceleration which contrast with periods of fairly constant acceleration.
Figure {\ref{fig:labeled_dives}} gives a more detailed view, showing a segment of the data containing six dives of at least two different types.
The plot shows one component of the acceleration, and from this we can see fine-scale behaviour within each dive which seems to indicate some fine scale short-term periodicity. 
We seek to understand both the coarse scale behaviour (e.g.~the dive type) and the fine-scale, within-dive behaviour. 
Such an analysis helps biologists to understand the foraging behaviour and feeding success of killer whales.

@@

From the plotted killer whale data in Figure {\ref{fig:labeled_dives}}, we see complexities within each dive that would be difficult to model as a realization of a simple stochastic process. 
In FDA, within curve structure is modelled by a generic smooth mean function and covariance function.
Alternatively, this structure is sometimes generated via random regression.
However, these classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in Figure 9.
We model the within-curve dependence by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence. Finally, for the fine-scale behaviour, we use summary information that reflects the application area's meaning for the curve.  Specifically, for the dive data, we use summaries of Fourier transforms to account for sudden and extreme ``wiggliness'' within some segments of a curve. The distribution of this ``wiggliness'' depends on the hidden state.

@@

Our work relies on work in the vast field of methodology for the analysis of animal movement
\citep{Hooten:2017}.
One of the most prevalent techniques in the literature of late is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \citep{Patterson:2017}. 
%
%Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another when conditioned on the underlying behavioral process.
%However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
%For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
%Several publications have dealt with this issue in the past, including the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
%The CarHMM in particular explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.
%while maintaining the structure needed to run the forward algorithm. 
%It also only adds one additional parameter per possible hidden state.

Animal movement modelers have recognized that several simultaneous behavioral processes may occur at different time scales. 
One approach to the analysis of this type of data is to use a hierarchical hidden Markov model, or HHMM \citep{Barajas:2017,Adam:2019}. 
HHMMs model the entire time series in question as a nested structure of hidden Markov models where each HMM corresponds to one behavioral process.

At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by traditional HMMs, CarHMMs, or HHMMs. Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}. 
%One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation. Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals. However, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.
One way to model periodic behavior in particular is to use signal processing techniques such as the Fourier transform on the raw data. 
The advantages of using Fourier analysis within an HMM has been recently demonstrated in the context of describing daily behavioral cycles of marine mammals \citep{Heerah:2017}.
In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017} and specifically fluking \citep{Shorter:2017} from accelerometer data. 
Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appears a promising simple approach to account for additional correlation in data that is cyclical in nature.
